Details of specific functions

### 2024-11-11 inst/gitbook/03_specific-tutorials.Rmd
Updated tutorials

## Using the first 10 dimensions
boot.matrix(BeckLee_mat50, dimensions = 10)
```

Of course, one could directly supply the subsets generated above (using `chrono.subsets` or `custom.subsets`) to this function.

@@

@@

```{r, eval = TRUE}
## Bootstrapping the observations (default)
boot_obs <- boot.matrix(data = crown_stem, boot.by = "rows")

## Bootstrapping the columns rather than the rows
boot_dim <- boot.matrix(data = crown_stem, boot.by = "columns")
```

In these two examples, the first one `boot_obs` bootstraps the rows as showed before (default behaviour).
But the second one, `boot_dim` bootstraps the dimensions.
That means that for each

```{r, eval = TRUE}
## Measuring disparity and summarising
summary(dispRity(boot_obs, metric = mean))
summary(dispRity(boot_dim, metric = mean))
```

### Bootstrapping with fine tuned probabilities

```{r, eval = TRUE}
boot.matrix(data = BeckLee_mat50)
```

@@

dist.data = TRUE

@@

boot.by = "dist"

>>>>

## Using the first 10 dimensions
boot.matrix(BeckLee_mat50, dimensions = 10)
```
Of course, one could directly supply the subsets generated above (using `chrono.subsets` or `custom.subsets`) to this function.

@@

### Bootstrapping with probabilities

@@

```{r, eval = TRUE}
## Bootstrapping the observations (default)
set.seed(1)
boot_obs <- boot.matrix(data = crown_stem, boot.by = "rows")

## Bootstrapping the columns rather than the rows
set.seed(1)
boot_dim <- boot.matrix(data = crown_stem, boot.by = "columns")
```

In these two examples, the first one `boot_obs` bootstraps the rows as showed before (default behaviour).
But the second one, `boot_dim` bootstraps the dimensions.
That means that for each bootstrap sample, the value calculated is actually obtained by reshuffling the dimensions (columns) rather than the observations (rows).

```{r, eval = TRUE}
## Measuring disparity and summarising
summary(dispRity(boot_obs, metric = sum))
summary(dispRity(boot_dim, metric = sum))
```

Note here how the observed sum is the same (no bootstrapping) but the bootstrapping distributions are quiet different even though the same seed was used.

@@

By default, the `dispRity` package does not treat any matrix as a distance matrix.
It will however try to guess whether your input data is a distance matrix or not.
This means that if you input a distance matrix, you might get a warning letting you know the input matrix might not be treated correctly (e.g. when bootstrapping or subsetting).
For the functions `dispRity`, `custom.subsets` and `chrono.subsets` you can simply toggle the option `dist.data = TRUE` to make sure you treat your input data as a distance matrix throughout your analysis.

```{r}
## Creating a distance matrix
distance_data <- as.matrix(dist(BeckLee_mat50))

## Measuring the diagonal of the distance matrix
dispRity(distance_data, metric = diag, dist.data = TRUE)
```

If you use a pipeline of any of these functions, you only need to specify it once and the data will be treated as a distance matrix throughout.

```{r}
## Creating a distance matrix
distance_data <- as.matrix(dist(BeckLee_mat50))

## Creating two subsets specifying that the data is a distance matrix
subsets <- custom.subsets(distance_data, group = list(c(1:5), c(6:10)), dist.data = TRUE)
## Measuring disparity treating the data as distance matrices
dispRity(subsets, metric = diag)

## Measuring disparity treating the data as a normal matrix (toggling the option to FALSE)
dispRity(subsets, metric = diag, dist.data = FALSE)
## Note that a warning appears but the function still runs
```

@@

The function `boot.matrix` also can deal with distance matrices by bootstrapping both rows and columns in a linked way (e.g. if a bootstrap pseudo-replicate draws the values 1, 2, and 5, it will select both columns 1, 2, and 5 and rows 1, 2, and 5 - keeping the distance structure of the data).
You can do that by using the `boot.by = "dist"` function that will bootstrap the data in a distance matrix fashion:

```{r}
## Measuring the diagonal of a bootstrapped matrix
boot.matrix(distance_data, boot.by = "dist")
```

Similarly to the `dispRity`, `custom.subsets` and `chrono.subsets` function above, the option to treat the input data as a distance matrix is recorded and recycled so there is no need to specify it each time.
